---
title: "Systematic CTN screening in NSCLC IMC data"
output: 
  rmarkdown::html_vignette:
    code_folding: "show"
vignette: >
  %\VignetteIndexEntry{NSCLC_IMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: 'Compiled: `r Sys.Date()`'
---

This tutorial shows an example of systematic screening of cell-type triad niches (CTN) in the non-small cell lung cancer (NSCLC) imaging mass cytometry (IMC) data using SCOPE. The dataset comes from [Cords et al. *Cancer Cell * 2024](https://www.cell.com/cancer-cell/fulltext/S1535-6108(23)00449-X) and is composed of 2,072 tissue samples from a cohort of 1,070 patients. 

First, we load the packages that will be used for the analysis and visualization.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, # Suppresses package startup messages
  warning = FALSE,  # Suppresses warning messages
  dpi = 300,
  out.width='100%', 
  fig.align = "center")
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r Load packages}
library(SCOPE.CTN)
library(data.table)
library(tidyverse)
library(cowplot)
library(ggdendro)
library(ggtext)
library(patchwork)
library(ComplexHeatmap)
dataset_name <- "NSCLC_IMC"
data_path <- paste0("vignette_data/", dataset_name, "/")   # !!! Change the directory
source(paste0(data_path, dataset_name, "_celltype_palette.R")) # cell type color palette
```

```{r Load input data for SCOPE, include=FALSE}
load(paste0(data_path, dataset_name, "_dat_in.RData"))
ncell_by_slide <- table(NSCLC_IMC_dat_in$ImageID)
slide_list = unique(names(ncell_by_slide)[which(ncell_by_slide>500)]) # 1984 images left
```

```{r Load results, include=FALSE}
load(paste0(data_path, dataset_name, "_CTN_jaccard_mat_hclust.RData"))
load(paste0(data_path, dataset_name, "_CTN_annotation_detailed.RData"))
load(paste0(data_path, dataset_name, "_CTN_merged_jaccard_mat_hclust.RData"))
load(paste0(data_path, dataset_name, "_CTN_merged_label.RData"))
load(paste0(data_path, dataset_name, "_CTN_merged_celltype_proportion.RData"))
load(paste0(data_path, dataset_name, "_Full_CTN_table_mgcv_merged.RData"))

CTN_list <- rownames(CTN_jaccard_mat_hclust$jaccard_mat_mgcv)
```

## 1. Prepare input data

The SingleCellExperiment object and clinical data for this dataset can be downloaded from [Zenodo](https://zenodo.org/records/7961844).

* `sce_all_annotated.rds` can be found in `SingleCellExperiment Objects.zip`.
* `clinical_data_ROI.csv` can be found in `comp_csv_files.zip`.

```{r load metadata, eval = FALSE}
library(SingleCellExperiment)
NSCLC_obj <- readRDS(paste0(data_path, "sce_all_annotated.rds"))
```

The input data frame for SCOPE should contain the following columns:

* `CellID`: cell id
* `ImageID`: sample id
* `X`: X coordinate
* `Y`: Y coordinate
* `Celltype`: cell type labels

```{r Formatting input, eval = FALSE}
# Extract cell metadata
NSCLC_IMC_dat_in = colData(NSCLC_obj)[,c("Tma_ac","CellID","cell_type","Center_X","Center_Y")]
NSCLC_IMC_dat_in <- as.data.frame(NSCLC_IMC_dat_in)

# Rename the columns
colnames(NSCLC_IMC_dat_in) <- c("ImageID","CellID","Celltype","X","Y")
NSCLC_IMC_dat_in$Celltype <- factor(NSCLC_IMC_dat_in$Celltype, NSCLC_IMC_celltype_levs)
```

We exclude the images with low cell counts (< 500) from further analysis.

```{r Image QC, eval=FALSE}
ncell_by_slide <- table(NSCLC_IMC_dat_in$ImageID)
slide_list = unique(names(ncell_by_slide)[which(ncell_by_slide>500)]) # 1984 images left
NSCLC_IMC_dat_in <- NSCLC_IMC_dat_in %>%
  filter(ImageID %in% slide_list)
save(NSCLC_IMC_dat_in, file = paste0(data_path, dataset_name, "_dat_in.RData"))
```


## 2. Create neighborhood cell proportion matrix

Next, we define the spatial neighborhood using use the k-nearest neighbors and compute the neighborhood cell type composition matrix. Here the number of nearest neighbors $k$ is set at default value 20.

In addition, we enforce a distance boundary for the nearest neighbors and exclude the cells falling outside this radius. The boundary is initially defined as the 99th percentile of all distances from focal cells to their nearest neighbors. If it falls outside a predefined range, we reset it to either the minimum (`min.radius`) or maximum threshold (`max.radius`).


```{r Compute neighborhood cell type proportion, eval=FALSE}
cell_neighbor_results_maxdist = pbmcapply::pbmclapply(
  1:length(slide_list), function(i) {
    Build_cell_neighbor_maxdist(
      dat_slide=dat_in[dat_in$ImageID==slide_list[i], ], knn_num = 20, 
      min.radius = 100, max.radius = 120, min.n.neighbors = 10,
      cell_types_available = levels(dat_in$Celltype),
      return.nn.ids = TRUE)},
  mc.cores = 5)


## cell type proportion within the neighborhood
cell_neighbor_table_maxdist = lapply(
  cell_neighbor_results_maxdist, "[[", "prop.df") %>% 
  do.call(rbind, .)
## cell counts by cell type within the neighborhood
cell_neighbor_counts = lapply(
  cell_neighbor_results_maxdist, "[[", "count.df") %>% 
  do.call(rbind, .)
## cell ids of the nearest neighbors
cell_neighbor_ids = lapply(
  cell_neighbor_results_maxdist, "[[", "neighbor.cell.ids") %>% 
  do.call(rbind, .)
## distance to nearest neighbors
dist.quantiles = lapply(
  cell_neighbor_results_maxdist, "[[", "dist.quantiles") %>% 
  do.call(rbind, .)

save(cell_neighbor_table_maxdist, 
     file = paste0(data_path, dataset_name, "_cell_neighbor_table_maxdist.RData"))
save(cell_neighbor_counts, 
     file = paste0(data_path, dataset_name, "_cell_neighbor_counts.RData"))
save(cell_neighbor_ids, 
     file = paste0(data_path, dataset_name, "_cell_neighbor_ids.RData"))
save(dist.quantiles, 
     file = paste0(data_path, dataset_name, "_dist.quantiles.RData"))
rm(cell_neighbor_results_maxdist)
```

```{r load neighborhood results}
load(paste0(data_path, dataset_name, "_cell_neighbor_table_maxdist.RData"))
load(paste0(data_path, dataset_name, "_dist.quantiles.RData"))
```

We can examine the distribution of 99th percentile of all distance to the nearest neighbors to select the threshold parameters min.radius and max.radius accordingly. 

```{r Examine distance distribution, out.width='50%', fig.width=3, fig.height=2.2}
dist.quantiles %>%
  as.data.frame() %>% 
  ggplot(aes(x = `99%`)) +
  geom_histogram(color = "white", fill = "skyblue", bins = 20) +
  labs(y = "99th percentile of Dist.\nto Nearest Neighbors", x = NULL) +
  theme_cowplot(font_size = 12) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
```


Cells that retain fewer than `min.n.neighbors` neighbors within the distance boundary will have their neighborhood cell-type proportions set to `NaN` and will be excluded from subsequent niche clustering. Here we set `min.n.neighbors` to 10.

```{r Examine loose cells, fig.width=8, fig.height=3}
cells_to_exclude = apply(cell_neighbor_table_maxdist, 1, function(x) {any(is.na(x))}) %>%
  as.data.frame() %>%
  `colnames<-`("is_na") %>%
  rownames_to_column("CellID") %>%
  left_join(NSCLC_IMC_dat_in[, c("CellID", "X", "Y", "ImageID")], by = "CellID")

# Number of cells to exclude
table(cells_to_exclude$is_na)

cells_to_exclude %>%
  group_by(ImageID) %>%
  mutate(n_NaN = sum(is_na)) %>%
  filter(n_NaN > 10) %>% ungroup %>%
  ggplot(aes(x = X, y = Y, color = is_na)) +
  geom_point(shape = 16, size= 0.1) +
  scale_color_manual(values = c("brown2", "grey90"), 
                     breaks = c(TRUE, FALSE), name = "Cells To\nExclude") +
  guides(color = guide_legend(override.aes = list(size = 3), ncol = 1)) +
  facet_wrap(~ImageID, scales = "free", ncol = 7) +
  theme_void() +
  coord_cartesian(expand = FALSE) +
  theme(legend.key.size = unit(2.5, "mm"),
        panel.background = element_blank(),
        aspect.ratio = 1)
```


## 3. Prefilter cell type triads using CLQ permutation test

SCOPE identifies niches characterized by cell-type triads. However, as the number of cell types grows, the combinations of three cell types can increase exponentially. Therefore, we utilize a colocation quotient (CLQ) permutation test to prioritize cell-type triads (CTs) that exhibit significant spatial proximity. 

Please refer to the [Prefiltering cell type triads using CLQ permutation test](Triad_prefilter_CLQ.html) vignette for details. 

```{r CLQ permutation test, eval = FALSE}
load(paste0(data_path, dataset_name, "_cell_neighbor_counts.RData"))
load(paste0(data_path, dataset_name, "_cell_neighbor_ids.RData"))

# Compute observed CLQ
CLQ_obs <- compute_CLQ_observed(
  cell_neighbor_counts = cell_neighbor_counts,
  dat_in = NSCLC_IMC_dat_in)

# Compute permutated CLQ by randomly shuffling cell type labels for 500 times
CLQ_perm <- compute_CLQ_permutated(
  cell_neighbor_ids = cell_neighbor_ids, 
  dat_in = NSCLC_IMC_dat_in,
  iter.num = 500, seed = 42, n_cores = 5)

# Compute permutation test p-value
CLQ_perm_pval <- CLQ_permutation_test(CLQ_obs = CLQ_obs, CLQ_perm = CLQ_perm)
```

Note that the CLQ permutation test can be time consuming. Here we load the precomputed results and use the Benjamini-Hochberg (BH) adjusted p-values to prioritize cell type triads where all constituent pairs exhibit significant CLQs (FDR < 0.05) in at least 40 tissue sections.

```{r load CLQ permutation test results}
load(paste0(data_path, dataset_name, "_CLQ_perm_pval_500iter.RData"))
head(CLQ_perm_pval, 5)
```

```{r filter triads}
# Exclude scarce (< 500 cells in total) & uninformative cell types ("Other") 
ncell_by_celltype <- table(NSCLC_IMC_dat_in$Celltype)
celltype_to_exclude <- c(names(which(ncell_by_celltype < 500)), "Other")
print(paste("Cell type(s) to exclude:", celltype_to_exclude))

# All combinations of the rest of the cell types
celltype_combs <- combn(setdiff(NSCLC_IMC_celltype_levs,
                                celltype_to_exclude), 3) %>%
  t() %>% as.data.frame()

# Match CLQ permutation test p-values
celltype_combs_w_pval <- celltype_combs %>%
  left_join(CLQ_perm_pval, by = c("V1" = "Celltype", "V2" = "Celltype_neighbor"), 
            relationship = "many-to-many") %>%
  left_join(CLQ_perm_pval, by = c("V1" = "Celltype_neighbor", "V3" = "Celltype", "ImageID"),
            suffix = c(".11", "")) %>%
  left_join(CLQ_perm_pval, by = c("V1" = "Celltype", "V3" = "Celltype_neighbor", "ImageID"),
            suffix = c(".12", "")) %>%
  left_join(CLQ_perm_pval, by = c("V1" = "Celltype_neighbor", "V3" = "Celltype", "ImageID"),
            suffix = c(".21", "")) %>%
  left_join(CLQ_perm_pval, by = c("V2" = "Celltype", "V3" = "Celltype_neighbor", "ImageID"),
            suffix = c(".22", ""))  %>%
  left_join(CLQ_perm_pval, by = c("V2" = "Celltype_neighbor", "V3" = "Celltype", "ImageID"),
            suffix = c(".31", ".32")) 

# Filter CTs where cell type triads where all constituent pairs exhibit significant CLQs (FDR < 0.05)
combination_table_filtered = celltype_combs_w_pval %>% ungroup %>%
  filter((p_perm.adj.11 < 0.05 | p_perm.adj.12 < 0.05) & 
         (p_perm.adj.21 < 0.05 | p_perm.adj.22 < 0.05) & 
         (p_perm.adj.31 < 0.05 | p_perm.adj.32 < 0.05)) %>%
  group_by(V1, V2, V3) %>%
  summarise(n = n()) %>%
  filter(n > 40) %>%
  select(-n)

print(paste("No of remaining CTs:", nrow(combination_table_filtered)))

# save(combination_table_filtered, 
#     file = paste0(data_path, dataset_name, "_combination_table_filtered.RData"))
```


```{r remove low density cells}
celltype_to_exclude <- cell_neighbor_table_maxdist[, unique(unlist(combination_table_filtered))] %>%
  apply(2, function(x) {sum(x > 0.3, na.rm = TRUE)}) %>% sort() %>%
  as.data.frame %>%
  `colnames<-`("n") %>%
  filter(n <= 50) %>%
  rownames()

if(length(celltype_to_exclude) > 1) {
  comb_to_run <- combination_table_filtered %>%
    mutate(combs = paste(V1, V2, V3, sep = "_")) %>%
    filter(!str_detect(combs, paste(celltype_to_exclude, collapse = "|"))) %>%
    select(-combs)
} else {
  comb_to_run <- combination_table_filtered
}

```


## 4. Cell type triad niche (CTN) detection using SCOPE

Next, we apply SCOPE for systematic screening of CTNs using the default configurations for the following parameters:

* `min.prop`: The minimum neighborhood cell-type proportion threshold. Proportions below this value are set to 0 (default: 0.3).
* `mgcv_df`: The dimension of the basis used for thin-plate regression spline smoothing of the driver cell-type proportions (default: 15; see `mgcv::s()` for details).

Here we set the number of clusters `clusternum` at 2. When setting `save_results = TRUE`, the clustering results of each CT will be saved as individual `.RData` files under `output_dir`.

```{r Run SCOPE, eval = FALSE}
# Load prefiltered CTs
load(paste0(data_path, dataset_name, "_combination_table_filtered.RData"))

celltype_to_exclude <- cell_neighbor_table_maxdist[, unique(unlist(combination_table_filtered))] %>%
  apply(2, function(x) {sum(x > 0.3, na.rm = TRUE)}) %>% sort() %>%
  as.data.frame %>%
  `colnames<-`("n") %>%
  filter(n <= 50) %>%
  rownames()


run_SCOPE(
  combination_table = combination_table_filtered,
  cell_neighbor_table = cell_neighbor_table_maxdist,
  dat_in = NSCLC_IMC_dat_in,
  save_results = TRUE,
  output_dir = paste0(data_path, "CTN_K_2"),
  clusternum = 2, min.prop = 0.3,
  linear = FALSE, mgcv_df = 15, iter.max = 100)
```

We gather the outputs and save them into one data frame.

```{r Collect SCOPE output, eval = FALSE}
Full_CTN_table_mgcv <- data.frame(dir = list.files(
  paste0(data_path, "_CTN_K_2"), ".RData", full.names = TRUE)) %>%
  mutate(group = row_number() %/% 30) %>%
  split(f = .$group) %>%
  lapply(pull, dir) %>%
  lapply(function(dirs) {
    pbmclapply(dirs, function(dir) {
      load(dir)
      return(CTN_df)
      CTN_df %>% select(CellID, CTN, label) %>%
        mutate(CTN = str_replace_all(CTN, "_CAF", ".CAF")) %>%
        mutate(CTN = str_replace_all(CTN, "hypoxic_tpCAF", "hypoxic.tpCAF"))
    }, mc.cores = 10) %>%
      do.call(rbind, .)  %>% `rownames<-`(NULL) %>%
      pivot_wider(names_from = "comb", values_from = "label")
})

Full_CTN_table_mgcv <- lapply(Full_CTN_table_mgcv, function(df) {df %>% select(-CellID)}) %>%
  `names<-`(NULL) %>%
  do.call(cbind, .) %>%
  cbind(data.frame(CellID = Full_CTN_table_mgcv[[1]]$CellID), .)

Full_CTN_table_mgcv <- temp %>%
  left_join(dat_in, by = "CellID") %>%
  mutate(Celltype = as.character(Celltype)) %>%
  mutate(Celltype = str_replace(as.character(Celltype), "_CAF", ".CAF")) %>%
  mutate(Celltype = str_replace(Celltype, "hypoxic_tpCAF", "hypoxic.tpCAF")) 
  select(all_of(colnames(NSCLC_IMC_dat_in)), everything()) %>%
  
save(Full_CTN_table_mgcv, file = paste0(
  data_path, dataset_name, "_Full_CTN_table_mgcv.RData"))
rm(temp)
```

## 5. Merging highly overlapped CTNs

Sometimes CTNs driven by two different triads can be highly similar. To merge highly overlapping CTNs, we calculate Jaccard distance between every pair of CTNs and perform hierarchical clustering to reduce redundancy. 

```{r Merge CTNs, eval = FALSE}
load(paste0(data_path, dataset_name, "_Full_CTN_table_mgcv.RData"))

# All cell type triads
CTN_list <- setdiff(colnames(Full_CTN_table_mgcv), colnames(NSCLC_IMC_dat_in))

CTN_jaccard_mat_hclust <- jaccard_dist_hclust(
  Full_CTN_table = Full_CTN_table_mgcv, 
  CTN_ls = CTN_list, num_cores = 5)

save(CTN_jaccard_mat_hclust, file = paste0(
     data_path, dataset_name, "_CTN_jaccard_mat_hclust.RData"))
```


Silhouette scores are utilized to determine the optimal number of clusters. The final cluster count is set to 77, balancing both the mean and median Silhouette scores.

```{r Silouhette score, fig.width = 4, fig.height = 3, out.width='50%'}
n_clusters = 77 # Final number of clusters
draw_silhoutte_score(CTN_jaccard_mat_hclust$hclust_silhouette,
                     num_clusters = n_clusters)
```

Below is a heatmap visualization of the Jaccard similarity indices (1-Jaccard index) and hierarchical clustering results. Black boxes outline final CTN groupings. Asterisks (*) and dots (.) indicate highly similar (Jaccard index > 0.95) and moderately similar (0.9 $\leq$ Jaccard index $\leq$ 0.95) CTN pairs, respectively.

```{r Jaccard similarity matrix heatmap, fig.width = 14, fig.height = 13}
htmap_jaccard_index <- draw_htmap_jaccard_index(
  jaccard_mat = CTN_jaccard_mat_hclust$jaccard_mat_mgcv,
  cl = CTN_jaccard_mat_hclust$cl,
  num_clusters = n_clusters)  
```

```{r Cut tree, eval=FALSE}

# Hierarchical clustering results
CTN_merged_label <- as.data.frame(cutree(CTN_jaccard_mat_hclust$cl, k = n_clusters)) %>%
  rownames_to_column("CTN") %>%
  `colnames<-`(c("CTN", "cluster")) %>%
  mutate(CTN = factor(CTN, rownames(jaccard_mat_mgcv_ordered))) %>%
  arrange(CTN) %>%
  mutate(cluster = factor(cluster, unique(.$cluster))) %>%
  mutate(cluster = factor(as.numeric(cluster))) %>%
  mutate(CTN = as.character(CTN))

# all(CTN_merged_label$CTN == rownames(jaccard_mat_mgcv_ordered))
save(CTN_merged_label, file= paste0(data_path, dataset_name, "_CTN_merged_label.RData"))
```

After merging highly overlapping initial CTNs using hierarchical clustering, the next step is to assign new names to these consolidated niches. The package uses the `CTN_annotate()` function to rename consolidated CTNs based on the most frequent core cell types within their original triads. If a tie occurs between cell types, the function automatically resolves it by identifying the cell type with the highest overall proportion within the consolidated CTN.

First, we compute the cell type proportion within the newly consolidated CTNs, which will be used as one of the inputs to the `CTN_annotate()` function. 


```{r CTN cell type proportion, eval = FALSE}
CTN_merged_celltype_proportion <- CTN_merged_label %>%
  split(f = .$cluster) %>%
  lapply(function(CTN_merged_label_sub) {
  Full_CTN_table_mgcv[, c("Celltype", CTN_merged_label_sub$CTN)] %>% 
      filter(if_any(everything(), ~ .x == "CTN")) %>%
      count(Celltype) }) %>%
  rbindlist(idcol = "cluster") %>%
  group_by(cluster) %>%
  mutate(prop = n/sum(n)) %>% select(-n)


save(CTN_merged_celltype_proportion, file = paste0(
     data_path, dataset_name, "_CTN_merged_celltype_proportion.RData"))
```

Additional inputs include a data frame containing the original CTN names and their hierarchical cluster labels, and a character vector defining the levels or ordering of the cell types. The `CTN_annotate` function returns a list of two data frames:

* `annotation`: A mapping data frame that matches the original CTN names and their hierarchical cluster labels to the final consolidated names
* `core_celltypes_summ`: A summary table focusing on clusters that contain at least 2 initial CTNs. It displays how many times each core cell type appeared in the original CTN names and tracks their relative proportions within the merged CTN.


```{r Annotate merged CTs}
CTN_merged_label_annotated <- CTN_annotate(
  CTN_cluster_label = CTN_merged_label, 
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion, 
  celltype_levs = names(NSCLC_IMC_palette))
```

If the top two most frequent cell types only appear twice in the original CTN names, the function flags the consolidated CTN with an asterisk (*). We suggest double-checking the flagged CTNs and manually renaming them if necessary.

Here we also manually renamed a few CTNs with high blood ("Blood"), lymphatic endothelial ("Lymphatic"), and vCAF proportions with "Vessel".

```{r Manually renaming CTNs, eval = FALSE}
head(CTN_merged_label_annotated$core_celltypes_summary, n = 15)

CTN_annotation_detailed <- CTN_merged_label_annotated$annotation %>%
  mutate(annotation_final = case_when(
    cluster == 12 ~ "Bcell_CD4_Vessel",
    cluster == 14 ~ "Bcell_Vessel_vCAF",
    cluster == 18 ~ "CD8_Vessel_vCAF",
    cluster == 33 ~ "T_Myeloid_vCAF",
    cluster == 36 ~ "CD8_Neutrophil_Vessel",
    cluster == 41 ~ "Neutrophil_Vessel_vCAF",
    cluster == 52 ~ "Neutrophil_Vessel_SMA.CAF",
    cluster == 61 ~ "Vessel_SMA.CAF_vCAF",
    TRUE ~ annotation))
# save(CTN_annotation_detailed, file = paste0(
#  data_path, dataset_name, "_CTN_annotation_detailed.RData"))
```


```{r Create full CTN label dataframe post merging, eval = FALSE}
Full_CTN_table_mgcv_merged <- CTN_annotation_detailed %>%
  split(f = .$cluster) %>%
  sapply(function(annot_df) {
    Full_CTN_table_mgcv[, c("CellID", annot_df$CTN)] %>%
      column_to_rownames("CellID") %>%
      apply(1, function(x) {ifelse(any(x == "CTN"), "CTN", "Unassigned")}) %>%
      as.data.frame() %>%
      `colnames<-`(annot_df$annotation_final[1])
  }, simplify = FALSE, USE.NAMES = FALSE) %>%
  do.call(cbind, .) %>%
  rownames_to_column("CellID")

Full_CTN_table_mgcv_merged <- Full_CTN_table_mgcv_merged %>%
  inner_join(NSCLC_IMC_dat_in, by = "CellID") %>%
  select(all_of(colnames(NSCLC_IMC_dat_in)), everything())

save(Full_CTN_table_mgcv_merged, 
     file = paste0(data_path, dataset_name, "_Full_CTN_table_mgcv_merged.RData"))
```

We can further perform hierarchical clustering on the Jaccard distance between the consolidated CTNs and grouping them into major categories.

```{r Compute jaccard index on merged CTNs, eval = FALSE}
load(paste0(data_path, dataset_name, "_Full_CTN_table_mgcv_merged.RData"))
CTN_merged_list <- setdiff(colnames(Full_CTN_table_mgcv_merged), colnames(NSCLC_IMC_dat_in))
CTN_merged_jaccard_mat_hclust <- jaccard_dist_hclust(
  Full_CTN_table = Full_CTN_table_mgcv_merged, 
  CTN_ls = CTN_merged_list, num_cores = 5)
save(CTN_merged_jaccard_mat_hclust, file = paste0(
  data_path, dataset_name, "_CTN_merged_jaccard_mat_hclust.RData"))
```

These categories can be annotated by examining the cell type abundance of the constituent CTNs.

```{r CTN dendrogram, fig.width=8, fig.height=8.5}
CTN_merged_celltype_proportion <- CTN_merged_celltype_proportion %>%
  left_join(distinct(CTN_annotation_detailed, cluster, annotation_final), by = "cluster") %>%
  rename("annotation" = "annotation_final")

CTN_merged_dend <- draw_CTN_dendro(
  cl = CTN_merged_jaccard_mat_hclust$cl, h = 0.6,
  dot_fontsize = 11, y_lim = c(NA, 1),
  x_expand = c(0.003, 0.003),
  celltype_palette = c(NSCLC_IMC_palette, "Vessel" = "#393B79", 
                       "CAF" = "#780116", "T" = "royalblue"),
  rename_celltype = NSCLC_IMC_rename_celltype, 
  show_text = FALSE)


p_CTN_merged_celltype_prop <- draw_CTN_celltype_prop(
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion, 
  CTN_dend = CTN_merged_dend, 
  celltype_palette = NSCLC_IMC_palette,
  rename_celltype = NSCLC_IMC_rename_celltype,
  facet_var = "clust", x_expand = c(0.05, 0.05))

p_CTN_merged_celltype_prop + CTN_merged_dend$p + 
  plot_layout(guides = "collect", widths = c(1, 0.4))
```

## 6. Identify clinically-relevant CTNs

Once we obtain the final CTN labels, we will next examine their prognostic value using the detailed patient clinical data provided by the original data source. The analysis is restricted to treatment-naive samples from patients with lung adenocarcinoma (LUAD) or lung squamous cell carcinoma (LUSC).

```{r Read patient clinical info}
# Image-level clinical data
clinical_data_LUAD_LUSC <- fread(paste0(data_path, "clinical_data_ROI.csv")) %>%
  mutate(DX_group = case_when(
    DX.name == "Adenocarcinoma" ~ "LUAD",
    DX.name == "Squamous cell carcinoma" ~ "LUSC",
    TRUE ~ "Other")) %>%
  mutate(Ev.DFS = ifelse(OS > DFS, 1, 0)) %>%
  rename("ImageID" = "Tma_ac") %>%
  filter(DX_group %in% c("LUAD", "LUSC")) %>%
  filter(NeoAdj == "NoNeoAdjuvantTherapy") 


table(clinical_data_LUAD_LUSC$DX_group)
```

To test whether the CTNs predict patient survival, we fit a Cox proportional hazards regression model using patient-level CTN presence as the primary predictor. For a given patient, a CTN is considered present if it contains 100 or more cells within a single tissue in at least one tissue section. 

While we conduct a simple univariate Cox regression here for demonstration purposes, users can easily extend this to a multivariate analysis by specifying confounding variables via the `confounder` argument.

```{r Cox regression on CTN presence, eval=FALSE}
# Overall survival
coxph_res_OS <- CTN_presence_coxph(
    Full_CTN_table = Full_CTN_table_mgcv_merged, 
    CTN_ls = CTN_merged_list, 
    clinical_df = clinical_data_LUAD_LUSC,
    time = "OS", event = "Ev.O",
    patient_id = "Patient_ID", confounder = NULL,
    min.n.patients = 3, min.n.cells = 100)

# Disease-free survival
coxph_res_DFS <- CTN_presence_coxph(
    Full_CTN_table = Full_CTN_table_mgcv_merged, 
    CTN_ls = CTN_merged_list, 
    clinical_df = clinical_data_LUAD_LUSC,
    time = "DFS", event = "Ev.DFS",
    patient_id = "Patient_ID", confounder = NULL,
    min.n.patients = 3, min.n.cells = 100)

save(coxph_res_OS, coxph_res_DFS, file = paste0(
     data_path, dataset_name, "_CTN_merged_coxph_res.RData"))
```

The following plots display the CTNs associated with the overall survival and disease-free survival of NSCLC patients, respectively.

```{r Visualize cox regression results on OS, fig.width = 6.5, fig.height = 3.5, out.width="70%"}
load(paste0(data_path, dataset_name, "_CTN_merged_coxph_res.RData"))
# Overall survival
draw_CTN_coxph_logHR(
  coxph_res = coxph_res_OS,
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion,
  alpha = 0.1, top = 10, 
  celltype_levs = names(NSCLC_IMC_palette), dot_fontsize = 14,
  celltype_palette = c(NSCLC_IMC_palette, "Vessel" = "#393B79", 
                       "CAF" = "#780116", "T" = "royalblue"),
  rename_celltype = NSCLC_IMC_rename_celltype,
  line_width = 0.4, y_text_size = NULL,
  width_ratio = c(0.45, 0.55)) 
```

```{r Visualize cox regression results on DFS, fig.width = 6, fig.height = 2.2, out.width="70%"}
# Disease-free survival
draw_CTN_coxph_logHR(
  coxph_res = coxph_res_DFS,
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion,
  alpha = 0.1, top = 10, 
  celltype_levs = names(NSCLC_IMC_palette), dot_fontsize = 14,
  celltype_palette = c(NSCLC_IMC_palette, "Vessel" = "#393B79", 
                       "CAF" = "#780116", "T" = "royalblue"),
  rename_celltype = NSCLC_IMC_rename_celltype,
  line_width = 0.6, y_text_size = NULL,
  width_ratio = c(0.45, 0.55))
```

Below are two example CTNs associated with the disease-free survival. One is $\mbox{B cell/CD4}^+\mbox{ T cell/CD8}^+\mbox{ T cell}$, which resembles the lymphocyte aggregates.

```{r Visualize Bcell_CD4_CD8, fig.width = 8, fig.height = 3}
p_CTN_celltype <- draw_CTN_celltype(
  Full_CTN_table = Full_CTN_table_mgcv_merged,
  CTN = "Bcell_CD4_CD8", 
  img_list = c("175B_117", "88A_119", "176C_1", "88C_17", "87B_135", "176C_8"),
  celltype_palette = NSCLC_IMC_palette, 
  rename_celltype = NSCLC_IMC_rename_celltype,
  scales = "free", aspect.ratio = 1, legend.ncol = 2)
p_CTN_celltype$p_celltype / p_CTN_celltype$p_CTN +
  plot_layout(guides = "collect") 
```
The other is blood endothelial $\mbox{(BEC)/CD4}^+\mbox{ T cell/CD8}^+\mbox{ T cell}$, which preferentialy co-localized with normoxic tumor cells compared to hypoxic tumor cells.

```{r Visualize CD4_CD8_Blood, fig.width = 8, fig.height = 3}
p_CTN_celltype <- draw_CTN_celltype(
  Full_CTN_table = Full_CTN_table_mgcv_merged,
  CTN = "CD4_CD8_Blood", 
  img_list = c("178C_85", "88B_125", "86A_38", "175A_27", "175A_31", "87B_73"),
  celltype_palette = NSCLC_IMC_palette, 
  rename_celltype = NSCLC_IMC_rename_celltype,
  scales = "free", aspect.ratio = 1, legend.ncol = 2)
p_CTN_celltype$p_celltype / p_CTN_celltype$p_CTN +
  plot_layout(guides = "collect") 
```

```{r session info}
sessionInfo()
```
