---
title: "Systematic CTN screening in HGSOC CosMx data"
date: 'Compiled: `r Sys.Date()`'
---

This tutorial demonstrates how to conduct a systematically screening of Celltype Triad Niches (CTNs) in a High-Grade Serous Tubo-Ovarian Cancer (HGSOC) CosMx Spatial Molecular Imaging (SMI) dataset using SCOPE. The dataset is from the Discovery cohort of [Yeh et al. *Nature Immunology* 2024](https://www.nature.com/articles/s41590-024-01943-5) and consists of 100 tissue samples from 58 patients. 

First, we load the packages that will be used for the analysis and visualization.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, # Suppresses package startup messages
  warning = FALSE,  # Suppresses warning messages
  dpi = 300,
  out.width='100%', 
  fig.align = "center")
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r Load packages}
library(SCOPE.CTN)
library(data.table)
library(tidyverse)
library(cowplot)
library(ggdendro)
library(ggtext)
library(patchwork)
library(ComplexHeatmap)
dataset_name <- "HGSOC_CosMx"
data_path <- paste0("vignette_data/", dataset_name, "/")   # !!! Change the directory
source(paste0(data_path, dataset_name, "_celltype_palette.R")) # cell type color palette
```

```{r Load input data for SCOPE, include=FALSE}
load(paste0(data_path, dataset_name, "_dat_in.RData"))
slide_list <- unique(HGSOC_CosMx_dat_in$ImageID) # 100 images left
```

```{r Load results, include=FALSE}
load(paste0(data_path, dataset_name, "_cell_neighbor_table_maxdist.RData"))
load(paste0(data_path, dataset_name, "_dist.quantiles.RData"))

load(paste0(data_path, dataset_name, "_CTN_jaccard_mat_hclust.RData"))
load(paste0(data_path, dataset_name, "_CTN_annotation_detailed.RData"))
load(paste0(data_path, dataset_name, "_CTN_merged_jaccard_mat_hclust.RData"))
load(paste0(data_path, dataset_name, "_CTN_merged_label.RData"))
load(paste0(data_path, dataset_name, "_CTN_merged_celltype_proportion.RData"))
load(paste0(data_path, dataset_name, "_Full_CTN_table_mgcv_merged.RData"))
 
CTN_list <- rownames(CTN_jaccard_mat_hclust$jaccard_mat_mgcv)
CTN_merged_list <- setdiff(colnames(Full_CTN_table_mgcv_merged),
                           colnames(HGSOC_CosMx_dat_in))
```

## 1. Prepare input data

The Seurat object of the Discovery cohort (`ST_Discovery_so.rds`) can be downloaded from [Zenodo](https://zenodo.org/records/12613839). Metadata of the tissue sections are included in [Supplementary Table 2b](https://static-content.springer.com/esm/art%3A10.1038%2Fs41590-024-01943-5/MediaObjects/41590_2024_1943_MOESM3_ESM.xlsx) (`41590_2024_1943_MOESM3_ESM.xlsx`).

```{r load seurat object, eval = FALSE}
library(Seurat)
ST_Discovery_so <- readRDS(paste0(data_path, "ST_Discovery_so.rds"))
```

The input data frame for SCOPE should contain the following columns:

* `CellID`: cell id
* `ImageID`: sample id
* `X`: X coordinate
* `Y`: Y coordinate
* `Celltype`: cell type labels

Some cell types in this dataset are labeled with a _LC suffix, indicating "Low Confidence". In this analysis, we will merge these with their high-confidence counterparts. Additionally, we will combine all the $\mbox{CD8}^-$ T cell population excluding the regulatory T cells (i.e., the double-negative T cells `CD4.T.cell.DN` and the $\mbox{CD4}^+$ T cells `CD4.T.cell`) into one category.

```{r Formatting input, eval = FALSE}
ST_Discovery_so$cell.subtypes_combined <- str_remove(ST_Discovery_so$cell.subtypes, "_LC$")
ST_Discovery_so$cell.subtypes_combined <- str_remove(ST_Discovery_so$cell.subtypes_combined, ".DN$")

ST_Discovery_so$CellID <- Cells(ST_Discovery_so)

HGSOC_CosMx_dat_in = ST_Discovery_so@meta.data[,c("samples","CellID","cell.subtypes_combined","x","y")]
colnames(HGSOC_CosMx_dat_in) <- c("ImageID","CellID","Celltype","X","Y")

HGSOC_CosMx_dat_in$Celltype <- factor(HGSOC_CosMx_dat_in$Celltype, names(HGSOC_CosMx_palette))
```

Since all 100 tissue sections exceed 500 cells, no samples are excluded from subsequent downstream analyses.

```{r Image QC, eval=FALSE}
ncell_by_slide <- table(HGSOC_CosMx_dat_in$ImageID)
all(ncell_by_slide > 500)

slide_list <- unique(HGSOC_CosMx_dat_in$ImageID) # 100 images left
save(HGSOC_CosMx_dat_in, file = paste0(data_path, dataset_name, "_dat_in.RData"))
```


## 2. Create neighborhood cell proportion matrix

Next, we define the spatial neighborhood using use the k-nearest neighbors and compute the neighborhood cell type composition matrix. Here the number of nearest neighbors $k$ is set at the default value 20.

In addition, we enforce a distance boundary for the nearest neighbors and exclude the cells falling outside this radius. The boundary is initially defined as the 99th percentile of all distances from focal cells to their nearest neighbors. If it falls outside a predefined range, we reset it to either the minimum (`min.radius`) or maximum threshold (`max.radius`).


```{r Compute neighborhood cell type proportion, eval=FALSE}
cell_neighbor_results_maxdist = pbmcapply::pbmclapply(
  1:length(slide_list), function(i) {
    Build_cell_neighbor_maxdist(
      dat_slide=HGSOC_CosMx_dat_in[HGSOC_CosMx_dat_in$ImageID==slide_list[i], ], 
      knn_num = 20, min.radius = 100, max.radius = 500, min.n.neighbors = 10,
      cell_types_available = levels(HGSOC_CosMx_dat_in$Celltype))},
  mc.cores = 5)

## cell type proportion within the neighborhood
cell_neighbor_table_maxdist = lapply(
  cell_neighbor_results_maxdist, "[[", "prop.df") %>% 
  do.call(rbind, .)
dist.quantiles = lapply(
  cell_neighbor_results_maxdist, "[[", "dist.quantiles") %>% 
  do.call(rbind, .)

save(cell_neighbor_table_maxdist, file = paste0(
  data_path, dataset_name, "_cell_neighbor_table_maxdist.RData"))
save(dist.quantiles, file = paste0(
  data_path, dataset_name, "_dist.quantiles.RData"))
```


We can examine the distribution of 99th percentile of all distance to the nearest neighbors to select the threshold parameters min.radius and max.radius accordingly. 

```{r Examine distance distribution, out.width='50%', fig.width=3, fig.height=2.2}
dist.quantiles %>%
  as.data.frame() %>% 
  ggplot(aes(x = `99%`)) +
  geom_histogram(color = "white", fill = "skyblue", bins = 20) +
  geom_vline(xintercept = 500, linetype = 2, color = "black") +
  labs(y = "99th percentile of Dist.\nto Nearest Neighbors", x = NULL) +
  theme_cowplot(font_size = 12) +
  theme(axis.text.x = element_text(hjust = 1, angle = 45))
```


Cells that retain fewer than `min.n.neighbors` neighbors within the distance boundary will have their neighborhood cell-type proportions set to `NaN` and will be excluded from subsequent niche clustering. Here we set `min.n.neighbors` to 10.

```{r Examine loose cells, fig.width=12, fig.height=5}
cells_to_exclude = apply(cell_neighbor_table_maxdist, 1, function(x) {any(is.na(x))}) %>%
  as.data.frame() %>%
  `colnames<-`("is_na") %>%
  rownames_to_column("CellID") %>%
  left_join(HGSOC_CosMx_dat_in[, c("CellID", "X", "Y", "ImageID")], by = "CellID")

# Number of cells to exclude
table(cells_to_exclude$is_na)

cells_to_exclude %>%
  group_by(ImageID) %>%
  mutate(n_NaN = sum(is_na)) %>%
  filter(n_NaN > 10) %>% ungroup %>%
  ggplot(aes(x = X, y = Y, color = is_na)) +
  geom_point(shape = 16, size= 0.1) +
  scale_color_manual(values = c("brown2", "grey90"), 
                     breaks = c(TRUE, FALSE), name = "Cells To\nExclude") +
  guides(color = guide_legend(override.aes = list(size = 3), ncol = 1)) +
  facet_wrap(~str_remove(ImageID, "SMI_"), ncol = 15) +
  theme_void() +
  coord_fixed(expand = FALSE) +
  theme(legend.key.size = unit(2.5, "mm"),
        panel.background = element_blank(),
        strip.clip = "off")
```


## 3. Cell type triad niche (CTN) detection using SCOPE

The HGSOC CosMx dataset has `r unique(HGSOC_CosMx_dat_in$Celltype)` available cell types, which can form only `r choose(length(unique(HGSOC_CosMx_dat_in$Celltype)), 3)` possible cell-type triads. Given the relatively small number of combinations, we will skip the colocation quotient (CLQ) permutation test step and instead focus solely on filtering out cell types that are less frequent within local neighborhoods. 

Specifically, we set the minimum neighborhood cell type proportion threshold at 0.15. The double-positive T cells (`T.cell.DP`) and mast cells (`Mast.cell`) have fewer than 100 cells that pass this threshold. We exclude the 10 triads that contain both of these cell types, leaving 210 cell-type triads to screen.

```{r remove low density cells}
celltype_to_exclude <- apply(cell_neighbor_table_maxdist, 2, 
                             function(x) {sum(x > 0.15, na.rm = TRUE)})
celltype_to_exclude <- names(celltype_to_exclude)[celltype_to_exclude < 100]
celltype_to_exclude

combination_table_all <- as.data.frame(t(combn(levels(HGSOC_CosMx_dat_in$Celltype), 3)))
has_all_targets <- rowSums(do.call(cbind, lapply(celltype_to_exclude, function(x) {
    rowSums(combination_table_all == x) > 0}))) == length(celltype_to_exclude)

combination_table_filtered <- combination_table_all[!has_all_targets, ]
paste("Number of cell-type triads to screen:", nrow(combination_table_filtered))
```

Next, we apply SCOPE for the systematic screening of CTNs. We set the the minimum neighborhood cell-type proportion threshold `min.prop` at 0.15 and utilize the default configurations for the rest of the parameters. For this analysis, we set the number of clusters (`clusternum`) to 2. By setting `save_results = TRUE`, the clustering results for each CTN are automatically saved as individual `.RData` files in the specified `output_dir.`

```{r Run SCOPE, eval = FALSE}
run_SCOPE(
  combination_table = combination_table_filtered[1:2, ],
  cell_neighbor_table = cell_neighbor_table_maxdist,
  dat_in = HGSOC_CosMx_dat_in,
  save_results = TRUE,
  output_dir = paste0(data_path, "CTN_K_2"),
  clusternum = 2, min.prop = 0.15,
  linear = FALSE, mgcv_df = 15, iter.max = 100)
```

We compile the individual outputs and merge them into a single data frame. The first five columns match the original input data, and the subsequent columns contain the assigned CTN labels.

```{r Collect SCOPE output, eval = FALSE}
Full_CTN_table_mgcv <- list.files(
  paste0(data_path, "CTN_K_2"), ".RData", full.names = TRUE) %>%
  pbmclapply(function(dir) {
    load(dir)
    return(CTN_table) 
  }, mc.cores = 5) %>%
  do.call(rbind, .)  %>%
  select(-label_new) %>%
  `rownames<-`(NULL) %>%
  pivot_wider(names_from = "CTN", values_from = "label") %>%
  left_join(HGSOC_CosMx_dat_in, by = "CellID") %>%
  select(all_of(colnames(HGSOC_CosMx_dat_in)), everything())

save(Full_CTN_table_mgcv, file = paste0(data_path, dataset_name, "_Full_CTN_table_mgcv.RData"))
```

## 4. Merging highly overlapped CTNs

Sometimes CTNs driven by two different triads can be highly similar. To merge highly overlapping CTNs, we calculate Jaccard distance between every pair of CTNs and perform hierarchical clustering to reduce redundancy. 

```{r Merge CTNs, eval = FALSE}
load(paste0(data_path, dataset_name, "_Full_CTN_table_mgcv.RData"))

# All cell type triads
CTN_list <- setdiff(colnames(Full_CTN_table_mgcv), colnames(HGSOC_CosMx_dat_in))

CTN_jaccard_mat_hclust <- jaccard_dist_hclust(
  Full_CTN_table = Full_CTN_table_mgcv, 
  CTN_ls = CTN_list, num_cores = 5)

save(CTN_jaccard_mat_hclust, file = paste0(
     data_path, dataset_name, "_CTN_jaccard_mat_hclust.RData"))
```


Silhouette scores are utilized to determine the optimal number of clusters. The final cluster count is set to 65, balancing both the mean and median Silhouette scores. As shown in the figure below, while 65 is not the optimal choice, it provides a trade-off to keep the total number of clusters manageable while still preserving as much cluster separation as possible.

```{r Silouhette score, fig.width = 4, fig.height = 3, out.width='50%'}
n_clusters = 65 # Final number of hierarchical clusters
draw_silhoutte_score(CTN_jaccard_mat_hclust$hclust_silhouette,
                     num_clusters = n_clusters)
```

Below is a heatmap visualization of the Jaccard similarity indices (1-Jaccard index) and hierarchical clustering results. Black boxes outline final CTN groupings. Asterisks (*) and dots (.) indicate highly similar (Jaccard index > 0.95) and moderately similar (0.9 $\leq$ Jaccard index $\leq$ 0.95) CTN pairs, respectively.

```{r Jaccard similarity matrix heatmap, fig.width = 14, fig.height = 13}
htmap_jaccard_index <- draw_htmap_jaccard_index(
  jaccard_mat = CTN_jaccard_mat_hclust$jaccard_mat_mgcv,
  cl = CTN_jaccard_mat_hclust$cl,
  num_clusters = n_clusters)  
```

```{r Cut tree, eval=FALSE}
# Hierarchical clustering results
htmap_row_order <- unlist(row_order(htmap_jaccard_index))
CTN_merged_label <- as.data.frame(cutree(CTN_jaccard_mat_hclust$cl, k = n_clusters)) %>%
  rownames_to_column("CTN") %>%
  `colnames<-`(c("CTN", "cluster")) %>%
  # keep the CTNs in the same order as the heatmap
  mutate(CTN = factor(CTN, rownames(
    CTN_jaccard_mat_hclust$jaccard_mat_mgcv)[htmap_row_order])) %>% 
  arrange(CTN) %>%
  mutate(cluster = factor(cluster, unique(.$cluster))) %>%
  mutate(cluster = factor(as.numeric(cluster))) %>%
  mutate(CTN = as.character(CTN))

save(CTN_merged_label, file= paste0(data_path, dataset_name, "_CTN_merged_label.RData"))
```

After merging highly overlapping initial CTNs using hierarchical clustering, the next step is to assign new names to these consolidated niches. The package uses the `CTN_annotate()` function to rename consolidated CTNs based on the most frequent core cell types within their original triads. If a tie occurs between cell types, the function automatically resolves it by identifying the cell type with the highest overall proportion within the consolidated CTN.

First, we compute the cell type proportion within the newly consolidated CTNs, which will be used as one of the inputs to the `CTN_annotate()` function. 


```{r CTN cell type proportion, eval = FALSE}
CTN_merged_celltype_proportion <- CTN_merged_label %>%
  split(f = .$cluster) %>%
  lapply(function(CTN_merged_label_sub) {
  Full_CTN_table_mgcv[, c("Celltype", CTN_merged_label_sub$CTN)] %>% 
      filter(if_any(everything(), ~ .x == "CTN")) %>%
      count(Celltype) }) %>%
  rbindlist(idcol = "cluster") %>%
  group_by(cluster) %>%
  mutate(prop = n/sum(n)) %>% select(-n)


save(CTN_merged_celltype_proportion, file = paste0(
     data_path, dataset_name, "_CTN_merged_celltype_proportion.RData"))
```

Additional inputs include a data frame containing the original CTN names and their hierarchical cluster labels, and a character vector defining the levels or ordering of the cell types. The `CTN_annotate` function returns a list of two data frames:

* `annotation`: A mapping data frame that matches the original CTN names and their hierarchical cluster labels to the final consolidated names
* `core_celltypes_summ`: A summary table focusing on clusters that contain at least 2 initial CTNs. It displays how many times each core cell type appeared in the original CTN names and tracks their relative proportions within the merged CTN.


```{r Annotate merged CTs, eval = FALSE}
CTN_merged_label_annotated <- CTN_annotate(
  CTN_cluster_label = CTN_merged_label, 
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion, 
  celltype_levs = names(HGSOC_CosMx_palette))
```

If the top two most frequent cell types only appear twice in the original CTN names, the function flags the consolidated CTN with an asterisk (*). Duplicated new CTN names are marked with a plus sign (+). We suggest double-checking the flagged CTNs and manually renaming them if necessary.

Here we also manually adjust a few merged CTNs where the constituent original CTNs are formed by a combination of four cell types.

```{r Manually renaming CTNs, eval = FALSE}
head(CTN_merged_label_annotated$core_celltypes_summary, n = 15)

CTN_annotation_detailed <- CTN_merged_label_annotated$annotation %>%
    mutate(annotation_final = case_when(
    cluster == 5 ~ "B.cell_NK.cell_Malignant_Endothelial",
    cluster == 6 ~ "B.cell_T.cell_NK.cell_Malignant",
    cluster == 9 ~ "B.cell_T.cell_Mast.cell_Malignant",
    cluster == 35 ~ "B.cell_T.cell_NK.cell_Monocyte",
    cluster == 46 ~ "T.cell_NK.cell_Mast.cell_Fibroblast",
    TRUE ~ annotation
  ))

save(CTN_annotation_detailed, file = paste0(
     data_path, dataset_name, "_CTN_annotation_detailed.RData"))
```

Given a newly consolidated CTN, any cell belonging to at least one of its constituent CTNs will be labeled as "CTN", while the remaining cells will be labeled as "Unassigned".

```{r Create full CTN label dataframe post merging, eval = FALSE}
Full_CTN_table_mgcv_merged <- CTN_annotation_detailed %>%
  split(f = .$cluster) %>%
  sapply(function(annot_df) {
    Full_CTN_table_mgcv[, c("CellID", annot_df$CTN)] %>%
      column_to_rownames("CellID") %>%
      apply(1, function(x) {ifelse(any(x == "CTN"), "CTN", "Unassigned")}) %>%
      as.data.frame() %>%
      `colnames<-`(annot_df$annotation_final[1])
  }, simplify = FALSE, USE.NAMES = FALSE) %>%
  do.call(cbind, .) %>%
  rownames_to_column("CellID")

Full_CTN_table_mgcv_merged <- Full_CTN_table_mgcv_merged %>%
  inner_join(HGSOC_CosMx_dat_in, by = "CellID") %>%
  select(all_of(colnames(HGSOC_CosMx_dat_in)), everything())

save(Full_CTN_table_mgcv_merged, 
     file = paste0(data_path, dataset_name, "_Full_CTN_table_mgcv_merged.RData"))
```

We can further perform hierarchical clustering on the Jaccard distance between the consolidated CTNs and grouping them into major categories.

```{r Compute jaccard index on merged CTNs, eval = FALSE}
load(paste0(data_path, dataset_name, "_Full_CTN_table_mgcv_merged.RData"))
CTN_merged_list <- setdiff(colnames(Full_CTN_table_mgcv_merged),
                           colnames(HGSOC_CosMx_dat_in))
CTN_merged_jaccard_mat_hclust <- jaccard_dist_hclust(
  Full_CTN_table = Full_CTN_table_mgcv_merged, 
  CTN_ls = CTN_merged_list, num_cores = 5)
save(CTN_merged_jaccard_mat_hclust, file = paste0(
  data_path, dataset_name, "_CTN_merged_jaccard_mat_hclust.RData"))
```

These categories can be annotated by examining the cell type composition of the constituent CTNs.

```{r CTN dendrogram, fig.width=8, fig.height=8}
CTN_annotation_merged <- CTN_annotation_detailed %>%
  # Grouping CTNs based on cell type composition  
  mutate(CTN_group = case_when(
    cluster %in% 1:11 ~ "Malignant",
    cluster %in% 12:18 ~ "Immune/Endothelial",
    cluster == 19 ~ "Bcell/Endothelial",
    cluster %in% 20:25 ~ "Bcell-rich",
    cluster %in% 26:28 ~ "Monocyte/Fibroblast",
    cluster %in% 29:37 ~ "Immune/Malignant",
    cluster %in% 38:39 ~ "Mixed",
    cluster %in% 40:52 ~ "Immune/Fibroblast",
    cluster %in% 53:65 ~ "Immune")) %>%
  distinct(cluster, annotation_final, CTN_group)%>%
  rename("annotation" = "annotation_final") 
  

CTN_merged_celltype_proportion_w_annot <- CTN_merged_celltype_proportion %>%
  left_join(CTN_annotation_merged[, 1:2], by = "cluster") 

CTN_merged_dend <- draw_CTN_dendro(
  cl = CTN_merged_jaccard_mat_hclust$cl, h = 0.7,
  annot_var = "annotation",
  rename_celltype = HGSOC_CosMx_rename_celltype, 
  CTN_annotation = CTN_annotation_merged,
  CTN_group_var = "CTN_group",
  CTN_group_palette = HGSOC_CosMx_CTN_group_palette,
  celltype_palette = c(HGSOC_CosMx_palette, "T.cell" = "dodgerblue3"),
  dot_fontsize = 11, y_lim = c(NA, 1),
  x_expand = c(0.003, 0.003),
  show_text = FALSE)


p_CTN_merged_celltype_prop <- draw_CTN_celltype_prop(
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion_w_annot, 
  CTN_dend = CTN_merged_dend, 
  celltype_palette = HGSOC_CosMx_palette,
  rename_celltype = HGSOC_CosMx_rename_celltype,
  facet_var = "CTN_group_label", 
  x_expand = c(0.05, 0.05),
  radius_range = c(1, 6))

p_CTN_merged_celltype_prop + CTN_merged_dend$p + 
  plot_layout(guides = "collect", widths = c(1, 0.4))
```

## 5. Identify clinically-relevant CTNs

After obtaining the final CTN labels, we will next examine their prognostic value using the detailed patient clinical data provided by the original data source. One patient with unknown cause of death will be excluded from survival analysis. 

```{r Read patient clinical info}
# Image-level clinical data
clinical_data <- readxl::read_xlsx(
  paste0(data_path, "41590_2024_1943_MOESM3_ESM.xlsx"),
  sheet = "Table 2b", skip = 1) %>%
  filter(dataset == "Discovery") %>%
  filter(outcome != "Dead (other)") %>%
  mutate(OS = as.numeric(fu_time1),               # Follow-up time post diagnosis
         Ev.O = case_when(outcome == "Alive" ~ 0, # Outcome (0 = alive, 1 = dead)
                          str_detect(outcome, "[Dd]ead") ~ 1,
                          is.na(outcome) ~ NA)) %>%
  mutate(age = as.numeric(age)) %>%
  mutate(ICB = ifelse(str_to_lower(immunotherapy) %in% c(
    "not reported (lost to follow-up)", 
    "on atezolizumab  vs placebo, stable",
    "no", "na"), "No", "Yes")) %>%
  select(ImageID = profile, Patient_ID = patients, sites_binary,
         OS, Ev.O, treatment, age, stage, ICB)

head(clinical_data)
```

To test whether the CTNs predict patient survival, we fit a Cox proportional hazards regression model using patient-level CTN presence as the primary predictor. For a given patient, a CTN is considered present if it contains 100 or more cells within a single tissue in at least one tissue section. 

We conduct both a univariate and a multivariate Cox regression analysis adjusting for age at diagnosis, tumor stage, and treatment history.

```{r Cox regression on CTN presence}
# Overall survival, univariate Cox regression
coxph_res_OS <- CTN_presence_coxph(
    Full_CTN_table = Full_CTN_table_mgcv_merged, 
    CTN_ls = CTN_merged_list, 
    clinical_df = clinical_data,
    time = "OS", event = "Ev.O",
    patient_id = "Patient_ID", confounder = NULL,
    min.n.patients = 3, min.n.cells = 100)

# Overall survival, multivariate Cox regression
coxph_res_OS_adjusted <- CTN_presence_coxph(
    Full_CTN_table = Full_CTN_table_mgcv_merged, 
    CTN_ls = CTN_merged_list, 
    clinical_df = clinical_data,
    time = "OS", event = "Ev.O",
    patient_id = "Patient_ID", 
    confounder = c("treatment", "age", "stage", "ICB"),
    min.n.patients = 3, min.n.cells = 100)
```

The following plots show the CTNs associated with the overall survival of HGSOC patients.

```{r Univariate cox regression on OS, fig.width = 7, fig.height = 4, out.width="70%"}
# Overall survival, univariate
draw_CTN_coxph_logHR(
  coxph_res = coxph_res_OS,
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion_w_annot,
  alpha = 1, top = 10, 
  celltype_levs = names(HGSOC_CosMx_palette), dot_fontsize = 14,
  celltype_palette = c(HGSOC_CosMx_palette, "T" = "dodgerblue3"),
  rename_celltype = HGSOC_CosMx_rename_celltype,
  line_width = 0.4, y_text_size = NULL,
  width_ratio = c(0.45, 0.55)) 
```

```{r Multivariate cox regression on OS, fig.width = 7, fig.height = 3.5, out.width="70%"}
# Overall survival, multivariate
draw_CTN_coxph_logHR(
  coxph_res = coxph_res_OS_adjusted,
  CTN_merged_celltype_prop = CTN_merged_celltype_proportion_w_annot,
  alpha = 1, top = 10, 
  celltype_levs = names(HGSOC_CosMx_palette), dot_fontsize = 14,
  celltype_palette = c(HGSOC_CosMx_palette, "T" = "dodgerblue3"),
  rename_celltype = HGSOC_CosMx_rename_celltype,
  line_width = 0.4, y_text_size = NULL,
  width_ratio = c(0.45, 0.55)) 
```

Below is the lymphocyte aggregate-like $\mbox{B cell/CD4}^+\mbox{ T cell/CD8}^+\mbox{ T cell}$ CTN that is associated with the overall survival.

```{r Visualize Bcell_CD4_CD8, fig.width = 6, fig.height = 2.6}
p_CTN_celltype <- draw_CTN_celltype(
  Full_CTN_table = Full_CTN_table_mgcv_merged,
  CTN = "B.cell_CD4.T.cell_NK.cell", 
  img_list = c("SMI_T14_F011", "SMI_T12_F001", "SMI_T11_F005", "SMI_T14_F006"),
  celltype_palette = HGSOC_CosMx_palette, 
  rename_celltype = HGSOC_CosMx_rename_celltype,
  scales = "fixed", legend.ncol = 1)
p_CTN_celltype$p_celltype / p_CTN_celltype$p_CTN +
  plot_layout(guides = "collect")
```

Below shows the $\mbox{CD8}^+\mbox{ T cell/Fibroblast/Malignant}$ CTN which separate the stromal and malignant compartments of the tissue.

```{r Visualize CD8T_Fibroblast_Malignant, fig.width = 8, fig.height = 2.6}
p_CTN_celltype <- draw_CTN_celltype(
  Full_CTN_table = Full_CTN_table_mgcv_merged,
  CTN = "CD8.T.cell_Fibroblast_Malignant", 
  img_list = c('SMI_T10_F001','SMI_T10_F015','SMI_T12_F001',
               'SMI_T12_F009','SMI_T12_F016','SMI_T13_F001'),
  celltype_palette = HGSOC_CosMx_palette, 
  rename_celltype = HGSOC_CosMx_rename_celltype,
  scales = "fixed", legend.ncol = 1)
p_CTN_celltype$p_celltype / p_CTN_celltype$p_CTN +
  plot_layout(guides = "collect")
```

```{r session info}
sessionInfo()
```
